Supplementary Tables

### 2019-09-22 Supplementary_mat_met.Rmd
additional files

# 1. Functions defined for ceRNA models and workflow of method
We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (Figure S1) :

@@

![Workflow for simulation of competing endogenous RNA regulations. Graph object in steps 2-6 is saved and updated continuously.](workflow1.png){#figs1}

@@

- 1. We have evaluated graph in the steady state conditions as followings:

```{r, message=FALSE, warning=FALSE, fig.cap= "Minimal Dataset in Steady-state", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")
```

- 2. We have obtained graph after change on Gene2 expression as following:

```{r, message=FALSE, warning=FALSE, fig.keep='all', fig.cap= "Gene2 Upregulation on Minimal Dataset", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

- 3. We have determined regulations after Gene2 upregulation:

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2,  fig.show='hold', fig.cap = "Sequential iteration of minsamp data. a) First response of system to Gene2 upregulation (2nd iteration). b) Spreading of perturbation on system (3th iteration)", fig.subcap= NULL}

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", Downregulation = "blue", title = "Regulations after Gene2 Upregulation", 2)

@@

We have made the same analysis in presence of interaction factors.

```{r, message=FALSE, warning=FALSE, fig.cap= "Minimal Dataset with interaction factors in Steady-state", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset with interaction factors in steady-state conditions")

@@

```{r, message=FALSE, warning=FALSE, fig.cap= " When Gene2 is upregulated on Minimal Dataset with interaction factors", fig.width=5, fig.height=4}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation with interaction factors")
```

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Sequential iteration of minsamp data with interaction factors", fig.subcap= c("First response of system to Gene2 upregulation (2nd iteration)", "Spreading of perturbation on system (3th iteration)")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation with interaction factors",2)
```

@@

Genes targeted by multiple miRNAs (referred to as "common target") are of special interest since they are subject to cooperative effect. Also, they perturb more than one neighborhood. In our small dataset, minsamp, Gene4 is regulated by two miRNAs. Let's simulate perturbation effects triggered by Gene4.

```{r, message=FALSE, warning=FALSE, out.width='50%', fig.asp=1, fig.ncol = 2, fig.cap = "Perturbation of Gene4 on minsamp data with interaction factors", fig.subcap= c("1st iteration after Gene4 upregulation", "2nd iteration after Gene4 perturbation")}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Common Gene-Gene4 Upregulation without interaction factors",2)
```

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We committed the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values at Supplementary tables S2-3)

[#TODO çok karmaşık komutlar için ayrı bir klasör hazırlanabilir. burayada link verilebilir: "experimentalmirnagene" tablosu repodaki X.R script ile elde edilmiştir" gibi]

@@

```{r, message= FALSE, warning=FALSE}
data("experimentalmirnagene")

@@

```{r}
data("E9GE_mirnagenenormal")
head(E9GE_mirnagenenormal)
```

@@

```{r}
data("E9GE_mirnagenetumor")
head(E9GE_mirnagenetumor)
```

@@

We have compared two datasets that are obtained for the tumor and normal tissue samples of same patient. We tried to change expression of a gene in normal tissue as the same level in the tumor tissue.

For this step, we have determined the changes of the gene expression in terms of fold change:

```{r}
E9GE_mirnagenetumor%>%
  dplyr::select(hgnc_symbol, GE_tumor)%>%
  dplyr::inner_join((E9GE_mirnagenenormal%>%dplyr::select(hgnc_symbol, GE_normal)), by ="hgnc_symbol")%>%
  dplyr::mutate(FC= GE_tumor/GE_normal)%>%
  distinct()%>%
  filter(FC>2.5,  FC<3.5)-> three_fold_change

# ABCC1 gene has 5827 read count in tumor tissue although 1420 in normal tissue (FC=4.10)
```

[#TODO SERPINE2]

Secondly, we have determined the most important nodes of network. We applied `find_node_perturbation` function. We only defined nodes that affect the other nodes more than 1.05 fold change with 10 iteration when they increase 3 fold.

```{r, include=FALSE, eval=FALSE}
as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  find_node_perturbation(how = 3, cycle = 10, fast = 5)%>%
  filter(!is.na(perturbation_efficiency))->perturbationofnetwork

perturbationofnetwork
```
This dataset, perturbationofnetwork, includes 382 genes of totally 423 nodes.

Result of this, we obtained common nodes of these two datasets (i.e perturbationofnetwork and three_fold_change) and selected a gene, SERPINE2.

```{r, eval=FALSE}
three_fold_change%>%
  inner_join(perturbationofnetwork, by = c("hgnc_symbol"="name"))

#Selected node is SERPINE2
```
## 2.5 Approach of Method into Combined Datasets

@@

```{r, fig.cap= "Percentage of affected nodes of each iteration for SERPINE2 Gene"}
as.data.frame(E9GE_mirnagenenormal)%>%
 priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
 update_how("SERPINE2",2.75) %>%
 simulate(150) %>%
 find_iteration(limit=1, plot= TRUE) #limit=1 describes the change that is not taken into account.
```

The node amount of changed gene on the system in terms of percentage were shown in above. As seen, firstly, the changed gene count increase. The system which contains the hundreds of miRNAs and thousands of genes can slowly gain the steady-state again. At first glance, it can be assumed that when all nodes in the system are reached, stable state will be provided. However, although all nodes are reached, the nodes competing with each other cause the edits to continue for a while.

@@

```{r}
as.data.frame(E9GE_mirnagenenormal)%>%
 priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
 update_how("SERPINE2",2.75) %>%
 simulate(100) %>%
 find_iteration(limit=1, plot= FALSE)

#31
```

@@

```{r}

as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  update_how("SERPINE2",2.75)%>%
  simulate(62)

@@

```{r}
as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  update_how("SERPINE2",2.75)%>%
  simulate(62)%>%
  as_tibble()%>%
  select(name, initial_count, count_current)->simulation_results
```

>>>>

# 1. Functions defined for ceRNA models and workflow of method
We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (Figure S1 in Supplementary Figures) :

@@

@@

- 1. We have evaluated graph in the steady state conditions as followings (Figure S2 in Supplementary Figures):

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)
```

- 2. We have obtained graph after change on Gene2 expression as following (Figure S3 in Supplementary Figures):

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)
```

- 3. We have determined regulations after Gene2 upregulation (Figure S4 in Supplementary Figures):

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  simulate(cycle= 2)

@@

We have made the same analysis in presence of interaction factors (Sequentially shown at Figure S5-7 in Supplementary Figures).

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)

@@

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)
```

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate(cycle=2)
```

@@

Genes targeted by multiple miRNAs (referred to as "common target") are of special interest since they are subject to cooperative effect. Also, they perturb more than one neighborhood. In our small dataset, minsamp, Gene4 is regulated by two miRNAs. Let's simulate perturbation effects triggered by Gene4 (Shown at Figure S8 in Supplementary Figures .

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate(cycle = 2)
```

@@

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We committed the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values at Supplementary files, Process on miRNA:target pairs dataset for process codes and Supplementary Tables for used numeric values)

@@

```{r, message= FALSE, warning=FALSE}

experimentalmirnagene<-readRDS("data/experimentalmirnagene.RDS")

@@

```{r}
#saveRDS(E9GE_mirnagenenormal, "data/E9GE_mirnagenenormal.rda")

E9GE_mirnagenenormal<-readRDS("data/E9GE_mirnagenenormal.rda")
head(E9GE_mirnagenenormal)
```

@@

```{r}
#saveRDS(E9GE_mirnagenetumor, "data/E9GE_mirnagenetumor.rda")

E9GE_mirnagenetumor<-readRDS("data/E9GE_mirnagenetumor.rda")
head(E9GE_mirnagenetumor)
```

@@

We have determined the most important nodes of network. We applied `find_node_perturbation` function. We only defined nodes that affect the other nodes more than 1.05 fold change with 10 iteration when they increase 3 fold.

```{r, include=FALSE, eval=FALSE}

as_data_frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  find_node_perturbation(how = 3, cycle = 10, fast = 5)%>%
  filter(!is.na(perturbation_efficiency))->perturbationofnetwork

perturbationofnetwork%>%
  filter(type== "Competing")%>%
  arrange(desc(perturbation_efficiency))
```
This dataset, perturbationofnetwork, includes totally 423 effective nodes. We selected SERPINE2 gene to perturb the network because it is the most efficient node gene in network.

## 2.5 Approach of Method into Combined Datasets

@@

The node amount of changed gene on the system in terms of percentage were shown at Figure S9 in Supplementary Figures. As seen, firstly, the changed gene count increase. The system which contains the hundreds of miRNAs and thousands of genes can slowly gain the steady-state again. At first glance, it can be assumed that when all nodes in the system are reached, stable state will be provided. However, although all nodes are reached, the nodes competing with each other cause the edits to continue for a while.

@@

```{r}
library(ceRNAnetsim)
as.data.frame(E9GE_mirnagenenormal)%>%
 priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
 update_how("SERPINE2",2.75)%>%
 simulate(50) %>%
 find_iteration(limit=0, plot= FALSE)

@@

```{r}
as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  update_how("SERPINE2",2.75)%>%
  simulate(32)

@@

```{r}
as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  update_how("SERPINE2",2.75)%>%
  simulate(32)%>%
  as_tibble()%>%
  dplyr::select(name, initial_count, count_current)->simulation_results
```

### 2019-09-22 Supplementary_tables.Rmd
additional files

|seed_type| Coefficient |
|-----------|------------|
| `5-mer`   | `0.05` |
| `5-mer_noncanonical`   | `0.04` |
| `6-mer`   | `0.07` |
| `6-mer_noncanonical`   | `0.05` |
| `6-merA1`   | `0.07` |
| `6-merA1_noncanonical`   | `0.05` |
| `7-mer`   | `0.23` |
| `7-mer_noncanonical`   | `0.19` |
| `7-merA1`   | `0.19` |
| `7-merA1_noncanonical`   | `0.16` |
| `7-mer-8m`   | `0.25` |
| `7-mer-8m_nonacnonical`   | `0.21` |
| `8-mer`   | `0.43` |
| `8-mer_noncanonical`   | `0.35` |
| `9-mer`   | `0.43` |
| `9-mer_noncanonical`   | `0.35` |
| `none`   | `0.01` |

Table: Efficiency factors for seed types.

|location of binding region on target| Coefficient |
|-----------|------------|
| `3'UTR`   | `0.84` |
| `CDS`   | `0.42` |
| `5'UTR`   | `0.01` |

Table: Efficiency factors for binding regions on targets

```{r, eval=FALSE, include=FALSE}
data("E9GE_mirnagenenormal")

xtable(head(E9GE_mirnagenenormal),caption = "Example of E9GE_mirnagenenormal dataset" )
```

|hgnc_symbol| miRNA_name | mirna_RPM | GE_normal | Energy | seed_type_effect | region_effect |
|-----------|------------|----------|----------------|---------|-----------|----------|
| CCNG1 | hsa-let-7a-5p | 111204.15 | 5245.00 | -25.10 | 0.05 | 0.84 |
| DICER1 | hsa-let-7a-5p | 111204.15 | 3285.00 | -24.40 | 0.43 | 0.42 | 
| SESN1 | hsa-let-7a-5p | 111204.15 | 1179.00 | -22.20 | 0.05 | 0.93 |
| NIPBL | hsa-let-7a-5p | 111204.15 | 4503.00 | -22.10 | 0.05 | 0.42 | 
| INTS12 | hsa-let-7a-5p | 111204.15 | 600.00 | -21.90 | 0.05 | 0.42 | 
| FNIP1 | hsa-let-7a-5p | 111204.15 | 1248.00 | -21.80 | 0.43 | 0.84 |

@@

>>>>

\begin{table}[ht]
\centering
\caption{Efficiency factors for seed types.}
\begin{tabular}{rlr}
  \hline
 & seed\_type & seed\_type\_effect \\ 
  \hline
1 & 6-mer\_noncanonical & 0.05 \\ 
  2 & 9-mer & 0.43 \\ 
  3 & 6-mer & 0.07 \\ 
  4 & 8-mer & 0.43 \\ 
  5 & 7-mer & 0.23 \\ 
  6 & none & 0.01 \\ 
  7 & 5-mer\_noncanonical & 0.04 \\ 
  8 & 5-mer & 0.05 \\ 
  9 & 6-merA1\_noncanonical & 0.05 \\ 
  10 & 7-mer-8m\_noncanonical & 0.21 \\ 
  11 & 7-mer-8m & 0.25 \\ 
  12 & 8-mer\_noncanonical & 0.35 \\ 
  13 & 7-merA1\_noncanonical & 0.16 \\ 
  14 & 7-merA1 & 0.19 \\ 
  15 & 6-merA1 & 0.07 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{ Efficiency factors for binding regions on targets}
\begin{tabular}{rlr}
  \hline
 & region & region\_effect \\ 
  \hline
1 & 3UTR & 0.84 \\ 
  2 & CDS & 0.42 \\ 
  3 & 3UTRCDS & 0.93 \\ 
  4 & 5UTR & 0.01 \\ 
  5 & 5UTRCDS & 0.42 \\ 
  6 &  & 0.01 \\ 
  7 & intron & 0.01 \\ 
  8 & CDS3UTR & 0.93 \\ 
  9 & CDS5UTR & 0.42 \\ 
  10 & exon\_unclassified & 0.20 \\ 
  11 & CDS3UTRintron & 0.93 \\ 
  12 & 3UTRintron & 0.84 \\ 
  13 & CDSintron & 0.42 \\ 
  14 & 5UTRintron & 0.01 \\ 
  15 & 5UTR3UTR & 0.93 \\ 
  16 & CDS5UTR3UTR & 0.93 \\ 
   \hline
\end{tabular}
\end{table}

|Hugo_Symbol| miRNA_name | mirna_RPM | GE_normal | energy | seed_type_effect | region_effect |
|-----------|------------|----------|----------------|---------|-----------|----------|
| ENAH | hsa-let-7a-5p | 111204.15 | 7540 | -22.70 | 0.01 | 0.00 |
| GALNT2 | hsa-let-7a-5p | 111204.15 | 2824 | -20.50 | 0.14 | 0.00 | 
| RLF | hsa-let-7a-5p | 111204.15 | 1144 | -18.00 | 0.03 | 0.20 |
| MAST2 | hsa-let-7a-5p | 111204.15 | 2640 | -22.60 | 0.24 | 0.00 | 
| DOCK1 | hsa-let-7a-5p | 111204.15 | 4826 | -20.50 | 0.01 | 0.00 | 
| ZBTB16 | hsa-let-7a-5p | 111204.15 | 315 | -24.30 | 0.14 | 0.00 |

@@

|Variebles| Structure | Means|
|-----------|------------|------------|
| **Node Variables**
| `name`  | character | node name |
| `type`  | character |Competing or miRNA |
| `node_id`  | numeric | in on graph object |
| `initial_count`   | numeric | Initial Expression value of node |
| `count_pre`   | numeric | Expression value of node at previous regulation |
| `count_current`   | numeric | Existing expression value of node |
| `changes_variable`   | character | Regulation of node (Up, down or steady) |
|-----------|------------|------------|
| **Edge Variables**
| `Competing name` | character | name of genes|
| `miRNA name` | character | name of miRNAs|
| `Competing expression` | numeric | Expression values of competing elements at steady-state |
| `miRNA expression` | numeric | Expression values of miRNA elements at steady-state |
| `energy` | numeric | coefficient of miRNA:target interactions (binding affinity)|
| `seed type` | numeric | coefficient of miRNA:target interactions (binding affinity)|
| `region` | numeric | coefficient of miRNA:target interactions (degradation efficiency)|
| `afff factor` | numeric| coefficient scaled and combined affinity factor |
| `degg factor` | numeric |coefficient scaled and combined degradation factor| 
| `comp_count_list` | list | list of competing expression for each iteration | 
| `comp_count` | numeric | _pre: competing expression at previous iteration; _current: competing expression at present iteration |
| `mirna_count_list` | list | list of miRNA expression for each iteration | 
| `mirna_count` | numeric | _pre: miRNA expression at previous iteration; _current: miRNA expression at present iteration |
| `effect` | numeric | _pre: total miRNA reppressive effect on individual target at previous iteration ; _current: miRNA reppressive effect on individual target at present iteration |
| `effect_list` | list | list of miRNA reppressive effect on individual target for each iteration |

Table: The context graph object during the process.
